\vspace{-.5cm} Lights: a generalized joint model for high-dimensional multivariate longitudinal data and censored durations \vspace{.5cm}

### 2021-09-01 paper.tex
update matrix formulation of survival density
go back to a dirac comb form for the baseline estimate

Let us denote 
\[ \theta = \big(\xi_0^\top \cdots \xi_{K-1}^\top, \beta_0^\top \cdots \beta_{K-1}^\top, \phi^\top, \text{vech}(D), \lambda_0, \gamma_0^\top \cdots \gamma_{K-1}^\top\big)^\top \in \R^\vartheta \]
the collection of the $\vartheta \in \N_+$ unknown parameters to estimate. Note that for ease of notations, we include $\lambda_0$ in $\theta$ although it is a function.
To write the log-likelihood $\ell_n(\theta)$ (rescaled by $n^{-1}$)  for samples in $\cD_n$, corresponding to the joint distribution of the time-to-event and longitudinal outcomes, let us make the following hypothesis.
\begin{assumption}
\label{indep-hyp-3}
Assume that both the random effects vector $b_i$ and the group membership account for the association between the longitudinal and event outcomes, that is
\begin{equation}
  \label{eq:ind-hyp}
  f(t_i, \delta_i, y_i | b_i, G_i = k ; \theta) = f(t_i, \delta_i| b_i, G_i = k ; \theta) f(y_i | b_i, G_i = k ; \theta)
\end{equation}
for all $i=1, \ldots, n$.
\end{assumption}
Assumption~\ref{indep-hyp-3} is a generalization of classical hypothesis used in SREMs and JLCMs~\citep{hickey2016joint}.
Then, one has
\begin{align}
	\label{eq:log-lik}
	\ell_n(\theta) &= \ell_n(\theta ; \cD_n) \nonumber \\
	&= n^{-1} \sum_{i=1}^n \log \sum_{k=0}^{K-1} \pi_{\xi_k}(x_i) f(y_i | G_i = k)\int_{\R^r} f(t_i, \delta_i| b_i, G_i = k ; \theta) f(b_i | y_i, G_i = k; \theta) \dd b_i,
\end{align}
where 
\[f(t_i, \delta_i| b_i, G_i = k ; \theta) = \big[\lambda(t_i|\cM_k(t_i), G_i = k)\big]^{\delta_i} \exp \Big\{-\int_0^{t_i} \lambda(s|\cM_k(s), G_i = k) \dd s \Big\}. \]

@@

\begin{equation}
  \label{eq:update_beta}
  {\beta_k}^{(w+1)}= \dfrac{\sum_{i=1}^n\hat \pi_{ik}^{\theta^{(w)}} \Big[U_i^\top I_{n_i} (y_i - V_i\hat \E_{\theta^{(w)}}[ b_i | t_i, \delta_i, y_i])\Big]}{\sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} U_i^\top I_{n_i} U_i}.
\end{equation}
Now, for the update of $\gamma_k^{(w)}$ for $k=0, \ldots, K-1$, we similarly denote
\begin{align*}
  Q^{(w)}_{n,k}(\gamma_k) = -n^{-1} \sum_{i=1}^n &\hat\pi_{ik}^{\theta^{(w)}} \Big[\delta_i \hat \E_{\theta^{(w)}}[ \gamma_k^\top \Psi_i(t_i) | t_i, \delta_i, y_i] \\
  & - \sum_{j=1}^{J-1} (t_{j+1}^u-t_j^u) \lambda_0^{(w)}(t_j^u) \hat \E_{\theta^{(w)}}[\exp \big\{\gamma_k^\top \Psi_i(t_j^u)\big\} | t_i, \delta_i, y_i] \ind{\{t_j^u < t_i\}} \Big]
\end{align*}
based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\gamma_k$. The integration over the survival process has been replaced with a finite sum over the process evaluated at the distinct event times $(t_1^u, \ldots, t_J^u)$ with $J \in~\N_+$, since $\lambda_0^{(w)}$ defined in~\eqref{eq:update_lambda_0} is piecewise-constant with jumps at observed event times. The update for $\gamma_k^{(w)}$ requires to solve the following minimization problem
\begin{equation}
  \label{eq:minimization-gamma_k}
  \gamma_k^{(w+1)} \in \argmin_{\gamma_k \in \R^{L\sA}} Q^{(w)}_{n,k}(\gamma_k) + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.
\end{equation}

@@

\vspace{.5cm}
Lemma~\ref{lemma:prox-sgl1} bellow states that one can easily compute the proximal operator of the sparse group lasso defined in Algorithm~\ref{alg:ISTA}.
\begin{lemma}
\label{lemma:prox-sgl1}
The proximal operator of the sparse group lasso can be expressed as the composition of the group lasso and the lasso proximal operators, that is
\normalfont
\begin{equation*}
  \text{prox}_{\text{sg} l_1,\tilde \eta, \zeta} = \text{prox}_{\zeta \tilde \eta \sum_{l=1}^L \norm{\cdot}_{2,l}} \circ \text{prox}_{\zeta (1 - \tilde \eta) \norm{\cdot}_1}.
\end{equation*}
\end{lemma}
The two proximal operators on the right hand side of Lemma~\ref{lemma:prox-sgl1} are both well known analytically~\citep{bach2011optimization} and tractable. We used the \texttt{Python} library \texttt{copt} for the implementation of proximal gradient descent~\citep{copt} and we propose in our package an implementation for the proximal operator of the sparse group lasso based on Lemma~\ref{lemma:prox-sgl1}.
A proof of Lemma~\ref{lemma:prox-sgl1} is proposed in Appendix~\ref{sec:proof-lemma-prox-sgl1}.
Hence, we update $\gamma_k^{(w)}$ with the following step
\begin{equation}
  \label{eq:gamma-update}
  \gamma_k^{(w+1)} = \text{ISTA}_{\text{sg} l_1}\big(\gamma_k^{(w)}, Q^{(w)}_{n,k}(\cdot), \tilde \eta, \zeta_{2,k}\big)
\end{equation}
Note that the warmstart initialization is left to the user's choice, as well as the ISTA acceleration~\citep{beck2009fast}, since warmstart can diminish acceleration effectiveness~\citep{tibshirani2010proximal}. The gradient of $Q_{n,k}^{(w)}$ is here given by
\begin{align}
  \label{eq:grad-Q}
  &\frac{\partial Q_{n,k}^{(w)}(\gamma_k)}{\partial \gamma_{k,a}^l} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ \delta_i  \hat \E_{\theta^{(w)}}[ \varphi_a(t_i, b_i^l) | t_i, \delta_i, y_i] \\ \nonumber
   &\qquad\qquad - \sum_{j=1}^{J-1} (t_{j+1}^u-t_j^u) \lambda_0^{(w)}(t_j^u) \hat \E_{\theta^{(w)}}[ \varphi_a(t_j^u, b_i^l) \exp \big\{\gamma_k^\top \Psi_i(t_j^u)\big\} | t_i, \delta_i, y_i] \ind{\{t_j^u < t_i\}} \Big]
\end{align}
for all $a = 1, \ldots, \mathcal{A}$ and $l = 1, \ldots, L$.

Then we consider a piecewise-constant estimator $\lambda_0^{(w)}$ for the baseline hazard with jumps at each uncensored observed event times. Hence, we treat the jump size as parameters to be estimated~\citep{klein1992semiparametric}, and the closed-form update obtained by cancelling the gradient is given by
\begin{equation}
  \label{eq:update_lambda_0}
  \lambda_0^{(w+1)}(t_j)= \dfrac{\sum_{i=1}^n \delta_i \ind{\{t_j=t_i\}}}{\sum_{i=1}^n \sum_{k=0}^{K-1} \hat \pi_{ik}^{\theta^{(w)}} (t_{j+1}^u-t_j^u) \hat \E_{\theta^{(w)}}\Big[\exp \big\{{\gamma_k^{(w+1)}}^\top \Psi_i(t_j)\big\} | t_i, \delta_i, y_i \Big] \ind{\{t_i \geq t_j\}} },
\end{equation}
for all $j=1, \ldots, n$, which is a Breslow like estimator~\citep{breslow1972contribution} adapted to our model.

@@

Let us introduce the following useful notations
\begin{equation}
  \label{eq:g-tilde}
  \left\{
    \begin{array}{ll}
    \tilde g^1&: b_i \mapsto b_i,\\
    \tilde g^2&: b_i \mapsto b_ib_i^\top, \\
    \tilde g^3_{s}&: b_i \mapsto \Psi_i(s), \\
    \tilde g^4_{s, \gamma_k}&: b_i \mapsto \exp \big\{\gamma_k^\top g^3_{s}(b_i) \big\}, \\
    \tilde g^5_{s, \gamma_k}&: b_i \mapsto g^3_{s}(b_i)^\top g^4_{s}(b_i).
    \end{array}
  \right. 
\end{equation}
We also denote 
\[ \cI_1 = 
\begin{bmatrix}
  \ind{\{t_{1} = t_1^u\}}  & \cdots & \ind{\{t_{1} = t_J^u\}}\\
  \vdots &  \ddots & \vdots \\
  \ind{\{t_{n} = t_1^u\}} & \cdots & \ind{\{t_{n} = t_J^u\}}
\end{bmatrix}
\in \R^{n \times J}
\quad \text{and} \quad 
\cI_2 = 
\begin{bmatrix}
  \ind{\{t_{1} \leq t_1^u\}} & \cdots & \ind{\{t_{1} \leq t_J^u\}}\\
  \vdots &  \ddots & \vdots \\
  \ind{\{t_{n} \leq t_1^u\}} & \cdots & \ind{\{t_{n} \leq t_J^u\}}
\end{bmatrix}
\in \R^{n \times J} \]
some matrices to compare each event times with the vector of distinct event times. Those matrices are widely used in practice to replace indicators. For instance
\begin{align*}
f(t_i, \delta_i| b_i, G_i = k ; \theta^{(w)}) &= \big[\lambda_0^{(w)}(t_i) \exp \big\{{\gamma_k^{(w)}}^\top\Psi_i(t_i) \big\} \big]^{\delta_i} \exp \Big( - \sum_{j=1}^{J-1} (t_{j+1}^u-t_j^u) \lambda_0^{(w)}(t_j^u) \exp \big\{{\gamma_k^{(w)}}^\top \Psi_i(t_j^u)\big\} \ind{\{t_j^u < t_i\}} \Big) \\
& = \big[\lambda_0^{(w)}(t_i) g^4_{t_i, \gamma_k}(b_i) \big]^{\delta_i} \exp \Big( - \sum_{j=1}^{J-1} (t_{j+1}^u-t_j^u) \lambda_0^{(w)}(t_j^u) g^4_{t_j^u, \gamma_k}(b_i) \ind{\{t_j^u < t_i\}} \Big) \\
f(T, \Delta| S^{(w)}, G = k ; \theta^{(w)}) &= \big\{ \underset{J}{\text{sum }} \big[g^4_{\gamma_k} \lambda_0^{(w)} \cI_1 \big] \big\}^{\Delta} \exp \Big( - \underset{J}{\text{sum }} \big[g^4_{\gamma_k} \lambda_0^{(w)} \Delta T\cI_2 \big] \Big)
\end{align*}
where $\Delta T = (t^u_2 - t^u_1  \cdots t^u_J - t^u_{J - 1}) \in \R^{J - 1}$.

>>>>

Let us denote 
\[ \theta = \big(\xi_0^\top \cdots \xi_{K-1}^\top, \beta_0^\top \cdots \beta_{K-1}^\top, \phi^\top, \text{vech}(D), \lambda_0^\top, \gamma_0^\top \cdots \gamma_{K-1}^\top\big)^\top \in \R^\vartheta \]
the collection of the $\vartheta \in \N_+$ unknown parameters to estimate. Note that we include $\lambda_0$ in $\theta$. Indeed, we will estimate the baseline using nonparametric maximum likelihood, as a function taking mass at each failure time. So in $\theta$, $\lambda_0$ represents the vector of these baseline hazard quantities, with a dimension equal to the number of unique failure times (denoted $J$ thereafter).
To write the log-likelihood $\ell_n(\theta)$ (rescaled by $n^{-1}$)  for samples in $\cD_n$, corresponding to the joint distribution of the time-to-event and longitudinal outcomes, let us make the following hypothesis.
\begin{assumption}
\label{indep-hyp-3}
Assume that both the random effects vector $b_i$ and the group membership account for the association between the longitudinal and event outcomes, that is
\begin{equation}
  \label{eq:ind-hyp}
  f(t_i, \delta_i, y_i | b_i, G_i = k ; \theta) = f(t_i, \delta_i| b_i, G_i = k ; \theta) f(y_i | b_i, G_i = k ; \theta)
\end{equation}
for all $i=1, \ldots, n$.
\end{assumption}
Assumption~\ref{indep-hyp-3} is a generalization of classical hypothesis used in SREMs and JLCMs~\citep{hickey2016joint}.
Then, one has
\begin{align}
	\label{eq:log-lik}
	\ell_n(\theta) &= \ell_n(\theta ; \cD_n) \nonumber \\
	&= n^{-1} \sum_{i=1}^n \log \sum_{k=0}^{K-1} \pi_{\xi_k}(x_i) f(y_i | G_i = k)\int_{\R^r} f(t_i, \delta_i| b_i, G_i = k ; \theta) f(b_i | y_i, G_i = k; \theta) \dd b_i,
\end{align}
where 
\[f(t_i, \delta_i| b_i, G_i = k ; \theta) = \big[\lambda(t_i|\cM_k(t_i), G_i = k)\big]^{\delta_i} \exp \Big\{-\int_0^{t_i} \lambda(s|\cM_k(s), G_i = k) \dd s \Big\}. \]

@@

\begin{equation}
  \label{eq:update_beta}
  {\beta_k}^{(w+1)}= \dfrac{\sum_{i=1}^n\hat \pi_{ik}^{\theta^{(w)}} \Big[U_i^\top I_{n_i} (y_i - V_i\hat \E_{\theta^{(w)}}[ b_i | t_i, \delta_i, y_i])\Big]}{\sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} U_i^\top I_{n_i} U_i}.
\end{equation}
Now, for the update of $\gamma_k^{(w)}$ for $k=0, \ldots, K-1$, we similarly denote
\begin{align*}
  Q^{(w)}_{n,k}(\gamma_k) = -n^{-1} \sum_{i=1}^n &\hat\pi_{ik}^{\theta^{(w)}} \Big[\delta_i \hat \E_{\theta^{(w)}}[ \gamma_k^\top \Psi_i(t_i) | t_i, \delta_i, y_i] \\
  & - \sum_{j=1}^J \lambda_0^{(w)}(t_j^u) \hat \E_{\theta^{(w)}}[\exp \big\{\gamma_k^\top \Psi_i(t_j^u)\big\} | t_i, \delta_i, y_i] \ind{\{t_j^u \leq t_i\}} \Big]
\end{align*}
based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\gamma_k$. The integration over the survival process has been replaced with a finite sum over the process evaluated at the distinct event times $(t_1^u, \ldots, t_J^u)$ with $J \in~\N_+$, since $\lambda_0^{(w)}$ defined in~\eqref{eq:update_lambda_0} is always zero except at observed event times. The update for $\gamma_k^{(w)}$ requires to solve the following minimization problem
\begin{equation}
  \label{eq:minimization-gamma_k}
  \gamma_k^{(w+1)} \in \argmin_{\gamma_k \in \R^{L\sA}} Q^{(w)}_{n,k}(\gamma_k) + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.
\end{equation}

@@

\vspace{.5cm}
Lemma~\ref{lemma:prox-sgl1} bellow states that one can easily compute the proximal operator of the sparse group lasso defined in Algorithm~\ref{alg:ISTA}.
\begin{lemma}
\label{lemma:prox-sgl1}
The proximal operator of the sparse group lasso can be expressed as the composition of the group lasso and the lasso proximal operators, that is
\normalfont
\begin{equation*}
  \text{prox}_{\text{sg} l_1,\tilde \eta, \zeta} = \text{prox}_{\zeta \tilde \eta \sum_{l=1}^L \norm{\cdot}_{2,l}} \circ \text{prox}_{\zeta (1 - \tilde \eta) \norm{\cdot}_1}.
\end{equation*}
\end{lemma}
The two proximal operators on the right hand side of Lemma~\ref{lemma:prox-sgl1} are both well known analytically~\citep{bach2011optimization} and tractable. We used the \texttt{Python} library \texttt{copt} for the implementation of proximal gradient descent~\citep{copt} and we propose in our package an implementation for the proximal operator of the sparse group lasso based on Lemma~\ref{lemma:prox-sgl1}.
A proof of Lemma~\ref{lemma:prox-sgl1} is proposed in Appendix~\ref{sec:proof-lemma-prox-sgl1}.
Hence, we update $\gamma_k^{(w)}$ with the following step
\begin{equation}
  \label{eq:gamma-update}
  \gamma_k^{(w+1)} = \text{ISTA}_{\text{sg} l_1}\big(\gamma_k^{(w)}, Q^{(w)}_{n,k}(\cdot), \tilde \eta, \zeta_{2,k}\big)
\end{equation}
Note that the warmstart initialization is left to the user's choice, as well as the ISTA acceleration~\citep{beck2009fast}, since warmstart can diminish acceleration effectiveness~\citep{tibshirani2010proximal}. The gradient of $Q_{n,k}^{(w)}$ is here given by
\begin{align}
  \label{eq:grad-Q}
  &\frac{\partial Q_{n,k}^{(w)}(\gamma_k)}{\partial \gamma_{k,a}^l} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ \delta_i  \hat \E_{\theta^{(w)}}[ \varphi_a(t_i, b_i^l) | t_i, \delta_i, y_i] \\ \nonumber
   &\qquad\qquad - \sum_{j=1}^J \lambda_0^{(w)}(t_j^u) \hat \E_{\theta^{(w)}}[ \varphi_a(t_j^u, b_i^l) \exp \big\{\gamma_k^\top \Psi_i(t_j^u)\big\} | t_i, \delta_i, y_i] \ind{\{t_j^u \leq t_i\}} \Big]
\end{align}
for all $a = 1, \ldots, \mathcal{A}$ and $l = 1, \ldots, L$.

For the update of $\lambda_0^{(w)}$, we treat the jump size at each uncensored observed event times as parameters to be estimated~\citep{klein1992semiparametric}, and the closed-form update obtained by cancelling the gradient is given by
\begin{equation}
  \label{eq:update_lambda_0}
  \lambda_0^{(w+1)}(t)= \dfrac{\sum_{i=1}^n \delta_i \ind{\{t=t_i\}}}{\sum_{i=1}^n \sum_{k=0}^{K-1} \hat \pi_{ik}^{\theta^{(w)}} \hat \E_{\theta^{(w)}}\Big[\exp \big\{{\gamma_k^{(w+1)}}^\top \Psi_i(t)\big\} | t_i, \delta_i, y_i \Big] \ind{\{t_i \geq t\}} },
\end{equation}
for all $t \geq 0$, which is a Breslow like estimator~\citep{breslow1972contribution} adapted to our model.

@@

Let us introduce the following useful notations
\begin{equation*}
  \left\{
    \begin{array}{ll}
    \tilde g^1&: b_i \mapsto b_i,\\
    \tilde g^2&: b_i \mapsto b_ib_i^\top, \\
    \tilde g^3_{s}&: b_i \mapsto \Psi_i(s), \\
    \tilde g^4_{s, \gamma_k}&: b_i \mapsto \exp \big\{\gamma_k^\top g^3_{s}(b_i) \big\}, \\
    \tilde g^5_{s, \gamma_k}&: b_i \mapsto g^3_{s}(b_i)^\top g^4_{s}(b_i).
    \end{array}
  \right. 
\end{equation*}
We also denote 
\[ \cI_1 = 
\begin{bmatrix}
  \ind{\{t_{1} = t_1^u\}}  & \cdots & \ind{\{t_{1} = t_J^u\}}\\
  \vdots &  \ddots & \vdots \\
  \ind{\{t_{n} = t_1^u\}} & \cdots & \ind{\{t_{n} = t_J^u\}}
\end{bmatrix}
\in \R^{n \times J}
\quad \text{and} \quad 
\cI_2 = 
\begin{bmatrix}
  \ind{\{t_{1} \leq t_1^u\}} & \cdots & \ind{\{t_{1} \leq t_J^u\}}\\
  \vdots &  \ddots & \vdots \\
  \ind{\{t_{n} \leq t_1^u\}} & \cdots & \ind{\{t_{n} \leq t_J^u\}}
\end{bmatrix}
\in \R^{n \times J} \]
some matrices to compare each event times with the vector of distinct event times. Those matrices are widely used in practice to replace indicators. For instance
\begin{align*}
f(t_i, \delta_i| b_i, G_i = k ; \theta^{(w)}) = &\big(\lambda_0^{(w)}(t_i) g^4_{t_i, \gamma_k^{(w)}}(b_i) \big)^{\delta_i} \\ 
&\times \exp \Big\{ - \sum_{j=1}^J \lambda_0^{(w)}(t_j^u) g^4_{t_j^u, \gamma_k^{(w)}}(b_i) \ind{\{t_j^u \leq t_i\}} \Big\}
\end{align*}
so that one can compute the tensor
\[ \underset{J}{\text{sum}} \big( \boldsymbol{g^4_{\gamma_k}} \boldsymbol{\lambda_0^{(w)}} \cI_1 \big)^{\boldsymbol{\delta}} \exp \big\{- \underset{J}{\text{sum}} \big(\boldsymbol{g^4_{\gamma_k}} \boldsymbol{\lambda_0^{(w)}} \cI_2 \big) \big\} \in \R^{n \times 2N} \]
where 
\begin{equation*}
  \left\{
    \begin{array}{ll}
    \boldsymbol{g^4_{\gamma_k}} \in \R^{2N \times J} \text{ with } \big[\boldsymbol{g^4_{\gamma_k}}\big]_{i, j} = g^4_{t_j^u, \gamma_k^{(w)}}(\breve{b}_i) , \\
    \boldsymbol{\lambda_0^{(w)}} = \big(\lambda_0^{(w)}(t_1^u), \ldots, \lambda_0^{(w)}(t_J^u) \big)^\top \in \R^J, \\
    \boldsymbol{\delta} = (\delta_1, \ldots, \delta_n)^\top \in \R^n.
    \end{array}
  \right. 
\end{equation*}
